### 2019-12-12 defesa/tex/chapter/e3p.tex
wip

The problem of finding every center and angle of rotation of a fixed shape ellipse which makes it have three points on its border is presented in this section. Even though its simple statement--it is short and uses only basic mathematical concepts--we were not able to find any work on it, or even on related problems. 
As a result, starting from scratch, we ended up trying a handful of approaches with most of them failing on the way. We try to give a review of some of those, and also make a case for the method we propose in terms of velocity of convergence and quality of the solutions that it finds.

We refer to this problem as \sigla{E3PNT}{Ellipse by Three points}, and an instance of it is given by three points $u, v, w \in \R^2$ and $E$, an ellipse with shape parameters $(a, b) \in \R^2_{>0}$, with $a > b$. A solution of E3PNT is a pair $(q, \theta) \in \R^2\times[0, \pi]$, such that $\{u, v, w\} \subset \tilde{E}(q, \theta)$. In other words, the goal is to develop a method to find every solution of E3PNT.

@@

To make it simpler, let us translate the system, so the point $u$ is at $(0,0)$. Then, we assume that the ellipse is actually axis-parallel and the points are the ones rotating. When an angle is found such that the three points lie on the border of the axis-parallel ellipse, a linear transformation can be applied to compress the x-axis by $\frac{b}{a}$, transforming the ellipse into a circle of radius $b$. This transformation can be seen on \autoref{fig:circumscribed-circle} where a solution of the E3PNT is transformed into a solution of the problem of finding a circumscribed circle of a triangle. 
This process can be parametrized by the angle of rotation of the points, as described by \autoref{eq:trpnts}, and because of the invertibility of linear transformations solutions for E3PNT can be obtained by reversing the transformations.

@@

Then, the problem to be solved is finding a circumscribed circle of the triangle formed by the points $(0, 0), \varphi(v, \theta)$ and $\varphi(w, \theta)$, such that the circle has radius $b$. As, for three non-colinear fixed points, there is always an unique circumscribed circle for the triangle formed by those three points, the only variable to be determined ends up being the angle of rotation $\theta$.

Let $A(\theta)$ be the area of the triangle formed by the points $(0, 0), \varphi(v, \theta)$ and $\varphi(w, \theta)$--note that the transformation does not preserve distance or area. Then, the radius $R$ of the circumscribed circle is given by \autoref{eq:circumscribed_circle} \cite[p.~189]{johnson1960}.

@@

Imposing the radius to be equal $b$ and squaring to eliminate the square roots present in the Euclidean distance, a function $\xi : [0, \pi) \mapsto \mathbb{R}_{>0}$ is defined by \autoref{eq:circumscribed_circle_b} in such a way that its zeros determine solutions to the E3PNT's instance. Two questions about $\xi(\theta)$ that arise are: is its set of roots finite? And, can they be found analytically?

@@

The method developed on \autoref{chapter:ellipses_n} iterates over every solution of E3PNT for every triplet of points, this is only possible if the size of this set of solutions is limited. Also, if this was not true, it would be very difficult to describe a method to get every solution which could be infinite.

It turns out that $\xi$ can be written as a real trigonometric polynomial of degree $6$ in the format given by \autoref{eq:trig_poly_2}, which implies that it can have up to $12$ distinct roots.
 To show that, just note that it is possible to write $\norm{\varphi(v, \theta)}^2$ and $A(\theta)^2$ in that form, as it can be seen on \autoref{eq:dd} and \autoref{eq:dd2}. It is also possible to see that the term of higher the degree of $\xi$ is the multiplication of the three squared distances. As $\norm{\varphi(v, \theta)}^2$ has degree $2$, the degree of $\xi$ is $6$.
\begin{align}\label{eq:dd}
	\norm{\varphi(v, \theta)}^2 = (v_x\frac{b}{a}\cos\theta + v_y\frac{b}{a}\sin\theta)^2 + (v_y\cos\theta - v_x\sin\theta)^2\\
	\label{eq:dd2} A(\theta)^2=\dfrac{1}{4}\det\left(
	\begin{array}{cc}
		v_x\frac{b}{a}\cos\theta + v_y\frac{b}{a}\sin\theta&v_y\cos\theta - v_x\sin\theta\\
		w_x\frac{b}{a}\cos\theta + w_y\frac{b}{a}\sin\theta&w_y\cos\theta - w_x\sin\theta
	\end{array}\right)^2
\end{align}

@@

On \autoref{chapter:definitions} a brief introduction is given on how to get the roots of a polynomial. For that reason, we discuss two ways of converting $\xi$ into a polynomial in this section. Symbolic computing was used to compute the polynomials, in practice an external Python library called SymPy was utilized (see \citeonline{sympy} for more details).

@@

The second approach is based on a idea published on \citeonline{boyd:2006} which uses the identities on \autoref{eq:complex_trig} to convert real trigonometric polynomials into univariate complex polynomials in order to obtain its roots using the companion matrix scheme.
Also, in \citeonline{weidner} it is said that computing the roots of a real trigonometric polynomial through this transformation does not yield loss of accuracy.

@@

It is possible to show that with that substitution and changing the variable to $z=e^{i\theta}$, we obtain the following function $g : \mathbb{S} \mapsto \mathbb{C}$, with $\mathbb{S}$ being the unit circle ($\mathbb{S} = \{z \in \mathbb{C} : |z|=1\}$):

\begin{equation}
g(z)=\sum_{k=0}^{12} c_k z^{k-6},
\end{equation}
for some $c_0, \dots, c_{12} \in \mathbb{C}$. Even though $g$ is not a polynomial, it is easy to obtain one from it. Firstly, the negative exponentials need to be shifted, this can be done by just multiplying $g$ by $z^6$. Secondly, the domain cannot be restricted to the unit circle, so we define $h : \mathbb{C} \mapsto \mathbb{C}$ as:
\begin{equation}
h(z) = z^6 g(z) = \sum_{k=0}^{12} c_k z^k.
\end{equation}

Then, by its definition is possible to see that every root of $g$ is also a root of $h$. Conversely, every root $\hat{z}$ of $h$ which is in $\mathbb{S}$--or $|\hat{z}|=1$--is also a root of $g$. Naturally, the angle of a root $\hat{z}$ of $g$ is a root of $\xi$ by Euler's Formula.

It is possible to make another reduction to obtain a degree-$6$ polynomial from $h$ whose roots form a superset of the roots of $g$. As it has been mentioned on this chapter, an ellipse is symmetric with respect to its own axis. This means $\theta$ and $\pi + \theta$ are equivalent angles of rotation for any ellipse, thus $\xi(\theta) = \xi(\pi + \theta)$\footnote{$\xi$ is only defined over $[0, \pi)$, so this equality is not actually valid.}. 
On \autoref{chapter:definitions}, it was stated that the angle of $z$ and $-z$ has the same symmetry with each other as an ellipse's angle of rotation:
\begin{equation*}
angle(-z) = \pi + angle(z).
\end{equation*}
From that, as $g(e^{i\theta})=\xi(\theta)$ for every $\theta \in [0, 2\pi)$, we conclude that $g(-z)=g(z)$. This implies that $h$ is, in fact an even polynomial, or that $h(-z) = h(z)$ is true for every $z\in\mathbb{C}$:
\begin{align}
h(-z) = (-z)^6g(-z) = z^6g(z).
\end{align}
Therefore, all the odd degree coefficients of $h$ are $0$ and we can define the $6$-degree polynomial $f : \mathbb{C} \mapsto \mathbb{C}$ with the substitution $y=z^2$:
\begin{equation}
f(y) = \sum_{k=0}^{6} c_{2k} y^k.
\end{equation}
Then from every root $\hat{y}$ of $f$, two roots of $h$ can be obtained: $\sqrt{\hat{y}}$ and $-\sqrt{\hat{y}}$. As the angle of $-\sqrt{\hat{y}}$ is not between $[0, pi)$ we can ignore it. Note that the the square root of $\hat{y}$ does not need to be calculated, as only the angles are needed and they can be obtained by:
\begin{equation}
angle(\sqrt{z}) = angle(z)/2.
\end{equation}
Finally, using the QR algorithm mentioned on \autoref{chapter:definitions} a $\bigO(n^3)$ algorithm, with $n=6$, can be constructed for E3P.

It is also worth mentioning that a pattern on the coefficients of $f$ was identified, and maybe for future work it can be used for further improvements. Analyzing the polynomials produced for several instances, the following seems to true:
\begin{equation}
c_k = \hat{c_{6-k}},
\end{equation}
for $k=0, \dots, 6$. For now, we do not have any ideas on how it could be proved. Nevertheless, this seems to lead somewhere interesting because this condition guarantees that $f$ is a self-reciprocal polynomial which implies that its roots will always come in pairs $(\hat{z}, 1/\hat{z})$.

>>>>

The problem of finding every center and angle of rotation of a fixed shape ellipse which makes it have three points on its border is presented in this section. Even though its simple statement--it is short and uses only basic mathematical concepts--we were not able to find any work on it, or even on related problems. 
As a result, starting from scratch, we ended up trying a handful of approaches with most of them failing on the way. We try to give a review of some of those going through the issues with the failing approaches, at the end we propose a $\bigO(1)$ algorithm that finds every solution of the problem and run some numerical experiments to show the precision of the found solutions.

We refer to this problem as \sigla{E3PNT}{Ellipse by Three points}, and an instance of it is given by three points $u, v, w \in \R^2$ and $E$, an ellipse with shape parameters $(a, b) \in \R^2_{>0}$, with $a > b$. A solution of E3PNT is a pair $(q, \theta) \in \R^2\times[0, \pi)$, such that $\{u, v, w\} \subset \tilde{E}(q, \theta)$.
The goal is to develop a method to find every solution of E3PNT.

@@

To make it simpler, let us translate the system, so the point $u$ is at $(0,0)$. Then, we assume that the ellipse is actually axis-parallel and the points are the ones rotating. When an angle is found such that the three points lie on the border of the axis-parallel ellipse, a linear transformation can be applied to compress the x-axis by $\frac{b}{a}$, transforming the ellipse into a circle of radius $b$. This transformation can be seen on \autoref{fig:circumscribed-circle} where a solution of the E3PNT is transformed into a solution of the problem of finding a circumscribed circle of a triangle. 
This process can be parametrized by the angle of rotation of the points, as described by \autoref{eq:trpnts}, and because of the invertibility of linear transformations, solutions for E3PNT can be obtained by reversing these transformations.

@@

After that transformation, the problem to be solved is finding a circumscribed circle of the triangle formed by the points $\{(0, 0), \varphi(v, \theta),\varphi(w, \theta)\}$, such that the circle has radius $b$. Therefore, the only unknown of this problem is the angle of rotation. This idea reduces the number of variables from $3$ to only $1$, however, this would only be useful if checking the existence of a circumscribed circle with radius $b$ is a convenient problem.

It turns out that for three non-colinear fixed points, there is always an unique circumscribed circle. Also, finding this unique circle can be done analytically, hence its radius can be easily computed and compared to $b$ when an angle is given.
Let $A(\theta)$ be the area of the triangle formed by the points $\{(0, 0), \varphi(v, \theta), \varphi(w, \theta)\}$--note that the transformation does not preserve distance or area. Then, the radius $R$ of the circumscribed circle is given by \autoref{eq:circumscribed_circle} \cite[p.~189]{johnson1960}.

@@

Imposing the radius to be equal $b$ and squaring to eliminate the square roots present in the Euclidean distance, a function $\xi : [0, 2\pi) \mapsto \mathbb{R}_{>0}$ is defined by \autoref{eq:circumscribed_circle_b} in such a way that its zeros determine solutions to the E3PNT's instance. Two questions about $\xi(\theta)$ that arise are: is its set of roots finite? And, can they be found analytically?

@@

One of the steps of the method developed on \autoref{chapter:ellipses_n} is to iterate over every solution of E3PNT for every triplet of points. Of course doing that is only possible for a finite number of solutions. Also, discovering a bound for that is essential for determining the efficiency of the algorithm.

Back on \autoref{chapter:definitions} real trigonometric polynomials were introduced. It was stated that a $n$-degree polynomial can have up to $2n$ distinct roots. It turns out that $\xi$ is a real trigonometric polynomial of degree $6$ and it can be written in the format given by \autoref{eq:trig_poly_2} which implies that it can have up to $12$ distinct roots.
 To show that, just note that it is possible to write $\norm{\varphi(v, \theta)}^2$ and $A(\theta)^2$ in that form, as it can be seen on \autoref{eq:dd} and \autoref{eq:dd2}. It is also possible to see that the term of higher the degree of $\xi$ is the multiplication of the three squared distances. As $\norm{\varphi(v, \theta)}^2$ has degree $2$, the degree of $\xi$ is $6$.
\begin{align}\label{eq:dd}
	\norm{\varphi(v, \theta)}^2 = (v_x\frac{b}{a}\cos\theta + v_y\frac{b}{a}\sin\theta)^2 + (v_y\cos\theta - v_x\sin\theta)^2\\
	\label{eq:dd2} A(\theta)^2=\dfrac{1}{4}\det\left(
	\begin{array}{cc}
		v_x\frac{b}{a}\cos\theta + v_y\frac{b}{a}\sin\theta&v_y\cos\theta - v_x\sin\theta\\
		w_x\frac{b}{a}\cos\theta + w_y\frac{b}{a}\sin\theta&w_y\cos\theta - w_x\sin\theta
	\end{array}\right)^2
\end{align}

@@

On \autoref{chapter:definitions} a brief introduction is given on how to get the roots of a polynomial. For that reason, we discuss two ways of converting $\xi$ into a polynomial in this section. Symbolic computing was used to compute the polynomials, the Python external library called SymPy was utilized (see \citeonline{sympy} for more details).

@@

The second approach is based on a idea published on \citeonline{boyd:2006} which uses the identities on \autoref{eq:complex_trig} to convert real trigonometric polynomials into univariate complex polynomials in order to obtain its roots using the companion matrix scheme. This approach is preferable as it preserves the numerical stability of the original real trigonometric polynomial--more details about this can be found in \citeonline{weidner} where it is said that computing the roots of a real trigonometric polynomial through this transformation does not yield loss of accuracy.

@@

It is possible to show that with that substitution and changing the variable to $z=e^{i\theta}$, we obtain the following function $g : \mathbb{S} \mapsto \mathbb{C}$, with $\mathbb{S}$ being the unit complex circle ($\mathbb{S} = \{z \in \mathbb{C} : |z|=1\}$):
\begin{equation}
g(z)=\sum_{k=0}^{12} c_k z^{k-6},
\end{equation}
for some $c_0, \dots, c_{12} \in \mathbb{C}$. As the equalities on \autoref{eq:complex_trig} are valid for any $\theta \in \R$, the function $g'(\theta)=g(e^{i\theta})$ and the function $\xi$ are equivalent since $g(e^{i\theta}) = \xi(\theta)$ for any $\theta \in [0, 2\pi)$.

Now there is only two things preventing $g$ from being a polynomial: firstly, $z$ shows up with negative exponents up to $-6$, secondly its domain is the unit circle, not the whole complex set $\mathbb{C}$. The first issue can be fixed by multiplying $g$ by $z^6$, note that this does not create further problems as $0\not\in \mathbb{S}$. The second issue is removed by extending the domain to $\mathbb{C}$, and considering only roots that lie on $\mathbb{S}$, notice also that no roots can go missing as $\mathbb{S} \subset \mathbb{C}$. Finally, the polynomial $h : \mathbb{C} \mapsto \mathbb{C}$ is defined as follows:
\begin{equation}
h(z) = z^6 g(z) = \sum_{k=0}^{12} c_k z^k.
\end{equation}

By its definition is possible to see that every root of $g$ is also a root of $h$, and conversely, every root $\hat{z}$ of $h$ which is in $\mathbb{S}$ is also a root of $g$. Naturally, given a root $\hat{z}$ of $g$, the angle of $\hat{z}$ is a root of $\xi$ by Euler's Formula used in the substitution $z=e^{i\theta}$.

It is possible to make another reduction to obtain a degree-$6$ polynomial, lowering in half the number of rows of the companion matrix utilized on obtaining the roots of $h$. 
As it has been mentioned before, an ellipse is symmetric with respect to its own axis. This means that $\theta$ and $\pi + \theta$ are equivalent angles of rotation for any ellipse, thus $\xi(\theta) = \xi(\pi + \theta)$.
On \autoref{chapter:definitions}, it was stated that the angle of $z$ and $-z$ has the same symmetry with each other as an ellipse's angle of rotation:
\begin{equation*}
angle(-z) = \pi + angle(z).
\end{equation*}
From that, as $g(e^{i\theta})=\xi(\theta)$ for every $\theta \in [0, 2\pi)$, we conclude that $g(-z)=g(z)$. This implies that $h$ is, in fact an even polynomial, or that $h(-z) = h(z)$ is true for every $z\in\mathbb{C}$:
\begin{align}
h(-z) = (-z)^6g(-z) = z^6g(z).
\end{align}
Therefore, all the odd degree coefficients of $h$ are $0$ and we can define the $6$-degree polynomial $f : \mathbb{C} \mapsto \mathbb{C}$ with the substitution $y=z^2$:
\begin{equation}
f(y) = \sum_{k=0}^{6} c_{2k} y^k.
\end{equation}
Then from every root $\hat{y}$ of $f$, two roots of $h$ can be obtained: $\sqrt{\hat{y}}$ and $-\sqrt{\hat{y}}$. As the angle of $-\sqrt{\hat{y}}$ is not between $[0, pi)$ we can ignore it. Note that the the square root of $\hat{y}$ does not need to be calculated, as only the angles are needed and they can be obtained by:
\begin{equation}
angle(\sqrt{z}) = angle(z)/2.
\end{equation}
Finally, using the QR algorithm mentioned on \autoref{chapter:definitions} a $\bigO(n^3)$ algorithm, with $n=6$, can be constructed for E3P.

It is also worth mentioning that a pattern on the coefficients of $f$ was identified, and maybe for future work it can be used for further improvements. Analyzing the polynomials produced for several instances, the following seems to be true:
\begin{equation}
c_k = \overline{c_{6-k}},
\end{equation}
for $k=0, \dots, 6$. For now, we do not have any ideas on how it could be proved. Nevertheless, this seems to lead somewhere interesting because this condition guarantees that $f$ is a self-reciprocal polynomial which implies that its roots will always come in pairs $(\hat{z}, 1/\hat{z})$.
